pipeComp, a general framework for the evaluation of computational pipelines, reveals performant single cell RNA-seq preprocessing tools

### 2020-05-07 bmc_article.tex

The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (Figure \ref{fig:figure2}A). In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics. Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments (or a desired subset), avoiding recomputing the same step twice and compiling evaluations (including run time) on the fly (see Supplementary methods for more detail). Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline. This is especially important because end-point metrics, such as the adjusted Rand index (ARI) \cite{HubertARI1985} for clustering, are not perfect. For example, although the meaning of an ARI score is independent of the number of true subpopulations \cite{steinleyProperties2004}, the number of clusters called is by far the most important determinant of the score: the farther it is from the actual number of subpopulations, the worse the ARI (Supplementary Figures 1-2). In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of clusters \cite{duoClustering2018}. While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown and tools that operate well in this optimal context might not necessarily be best overall. Clustering results are also very sensitive and might not always capture improvements in earlier steps of the pipeline. In addition, we wanted to assess whether the effect of a given parameter alteration is robust to changes in the rest of the pipeline. We therefore monitored several complementary metrics across multiple steps of the process. We proceeded in a step-wise fashion, first testing a large variety of parameters at the early steps of the pipeline along with only a set of mainstream options downstream, then selecting main alternatives and proceeding to a more detailed benchmark of the next step (Figure \ref{fig:figure2}B).

@@

While most methods accurately identified the doublets in the 3 cell lines dataset (mixology10x3cl), the other two datasets proved more difficult (Figure \ref{fig:figure3}A). \textit{scDblFinder} achieved comparable or better accuracy than top alternatives while being the fastest method (Figure \ref{fig:figure3}B). Across datasets, cells called as doublets tended to be classified in the wrong cluster more often than other cells (Figure \ref{fig:figure3}C). We also found that \textit{scDblFinder} improved the accuracy of the clustering across all benchmark datasets even when, by design, the data contained no heterotypic doublet (Figure \ref{fig:figure4}).

@@

We next investigated the impact of filtering according to six methods with varying stringency; three methods based on deviations to MADs with increasing levels of stringency (named \textit{lenient}, \textit{default}, \textit{stringent}, \textit{very stringent}) and two methods based on \textit{scater}'s \textit{runPCA} using all (\textit{pca}) or a restricted number of covariates (\textit{pca2}, see Methods for more details). An important risk of excluding cells on the basis of their distance from the whole distribution of cells on some properties (e.g., library size) is that these properties tend to have different distributions across subpopulations. As a result, thresholds in terms of number of MADs from the whole distribution can lead to strong biases against certain subpopulations (Figure \ref{fig:figure4}A). We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \ref{fig:figure4}B and Supplementary Figure 10). Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here. We therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm \cite{PapadimitriouHu1998} and monitored their mean F1 score. A first observation was that, although the stringent filtering tended to be associated with an increase in accuracy, it could also become deleterious. Most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias (Figure \ref{fig:figure4}B). Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) rather than on whole dataset resulted in nearly no cells being filtered out. This suggests that filtering on the global population tends to discard cells of subpopulations with more extreme properties (e.g. high library size), rather than low-quality cells, and that more stringent filtering may only remove the cells of these subpopulations. On the opposite, lenient or lack of filtering leads to no/ little impact on subpopulations. This comes at the cost that lenient filtering performs worse or equally well than other methods among all datasets. However, a mild filtering ('default') often leads to higher clustering performance (e.g. simMix and Zhengmix4 datasets) or to very little loss (e.g. Koh and Zhengmix8eq) and always retain most of the cells in each subpopulation. This suggests that a mild ('default') filtering is a very good trade-off between clustering accuracy and subpopulations preservation.  Finally, the doublet removal step in conjunction with filtering  resulted in a net decrease in the proportion of excluded cells while retaining or improving accuracy among some 10x datasets (mixology, Zheng datasets). On the Smart-Seq and simulated datasets however, the removal of doublets had little to no impact on any of the two measures. Overall, we therefore recommend the use of doublet removal followed by a `default' filtering (see Methods) or a similar approach.

@@

An important motivation for the development of \textit{SCTransform} was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an inadequate normalization \cite{hafemeisterSCtransform2019}. However, as library size tends to vary across subpopulations, part of this effect can simply reflect biological differences. We therefore assessed to what extent the first principal component still retained a correlation with the library size and the number of detected features, removing the confounding covariation with the subpopulations (Figure \ref{fig:figure5}A and Supplementary Figure 12). Scaling tended to remove much of the correlation with these features, and while most methods were able to remove most of the effect, the correlation was lowest with \textit{SCTransform}. Surprisingly, however, regressing out the covariates tended to increase the association with it.

We further evaluated normalization methods by investigating their impact on the separability of the subpopulations (Figure \ref{fig:figure5}B-C). Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width \cite{RousseeuwSil1987} and mutual information (MI). We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations. In view of the subpopulation silhouette and ARI at true number of clusters, it seems that all methods perform worse on the three Zhengmix datasets, that are the datasets with the highest number of cells and the smaller feature detection rate. Scaling tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones and was generally, but not always, beneficial on the accuracy of the final clustering. Regressing out covariates systematically gave equal  or poorer performance on MI, ARI and ARI at true number of clusters, with a notable negative impact when \textit{Seurat}'s regressing was applied to the Koh and Kumar datasets. It is possible that the difference between \textit{Seurat} and \textit{SCTransform}'s regression originates from the use of Pearson's residuals performed during the second round regression of \textit{SCTransform}, which are then independent of sequencing depth. \textit{SCTransform} systematically outperformed other methods and, even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).\textit{scVI} (Supplementary Figure 13) performed poorly on most datasets in terms of both the subpopulation silhouette and clustering accuracy. For time and resource cost reasons, \textit{scVI} was run on a reduced set of pipelines and compared to \textit{SCTransform} (more details in the Methods section).

@@

Reasoning that a selection method should ideally select genes whose variability is higher between subpopulations than within, we first assessed to what extent each method selected genes with a high proportion of variance or deviance explained by the (real) subpopulations. As the proportion of variability in a gene attributable to subpopulations can be measured in various ways, we first compared three approaches: ANOVA on log-normalized count, ANOVA on \textit{SCTransform} normalization, and deviance explained. The ANOVAs performed on a standard \textit{Seurat} normalization and on \textit{SCTransform} data were highly correlated (Pearson correlation >0.9, Supplementary Figure 17A). These estimates were also in good agreement with the deviance explained, although lowly-expressed genes could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 17B-D). We therefore compared the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 18-19). We first focused on the first 1000 genes to highlight the differences between methods, although a higher number of selected genes decreased the differences between methods (Supplementary Figures 18-20). The standardized measures of variability were systematically worse than their non-standardized counterparts in selecting genes with a high proportion of variance explained by subpopulation. Regarding the percentage of deviance explained however, the standardized measures were often superior (Figure \ref{fig:figure6}A and Supplementary Figures 18-19). \textit{Deviance} proved the method of choice to prioritize genes with a high variance explained by subpopulations (with mere expression level proving surprisingly performant) but did not perform so well to select genes with a high deviance explained.

We next evaluated how the use of different feature selection methods affected the clustering accuracy (Figure \ref{fig:figure6}B). To validate the previous assay on the proportion of variance/deviance explained by real populations, we included genes that maximized these two latter measures. Interestingly, while these selections were on average the top-ranking methods, they were not systematically best for all datasets. The ARI at true number of clusters was near perfect (>0.94) among the smallest datasets but among bigger datasets (simMix and ZhengMix) the performance of the methods was lower and more variable (Figure \ref{fig:figure6}B). Non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics. In general, we found deviance and unstandardized estimates of variance to provide the best results across datasets and normalization methods. Increasing the number of features selected also systematically led to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Supplementary Figure 20).

@@

We compared the various estimates of dimensionality in their ability to retrieve the intrinsic number of dimensions in a dataset, based on Seurat's weighted PCA space. As a first approximation of the true dimensionality, we computed the variance in each principal component that was explained by the subpopulations, which sharply decreased after the first few components in most datasets (Figure \ref{fig:figure8}A). Figure \ref{fig:figure8}B shows the difference between the dimension estimates of the above methods and that based on the subpopulations (i.e. from Figure \ref{fig:figure8}A). Of note, the methods differ widely in view of their computing time (Figure \ref{fig:figure8}B) and we saw no relationship between the accuracy of the estimates and the complexity of the method. Reasoning that over-estimating the number of dimensions is less problematic than under-estimating it, we kept the former methods for a full analysis of their impact on clustering (Figure \ref{fig:figure8}C-D), when combined with \textit{SCTransform} or the standard \textit{Seurat} normalization. Although most methods performed well on the various clustering measures, the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}, using the 20 nearest neighbors, although estimates were relatively stable across decent values of \textit{k}; see Supplementary Figure 23) provided the dimensionality estimate that best separated the subpopulations (Figure \ref{fig:figure8}C) and resulted in the best clustering accuracy (Figure \ref{fig:figure8}D). This method systematically selected many more components than were associated with the subpopulations, suggesting that although these additional components appear individually uninformative, in combination they nevertheless contribute to classification. An implication is that the commonly-used elbow method most likely underestimates dimensionality.

@@

Since ARI is dominated by differences in the number of clusters (Supplementary Figures 1-2) and no single metric is perfect, we diversified them (Figure \ref{fig:figure9}). MI has the virtue of not decreasing when a true subpopulation is split into two clusters, which is arguably less problematic (and might well reflect unknown biological subgroups), but as a consequence it can be biased towards methods producing higher resolution clustering (Supplementary Figure 2). Similarly, precision per true subpopulation is considerably more robust to differences in the number of clusters. We also tracked the mean F1 score and the ARI at the true number of clusters.

The MI score and minimum precision, which are largely independent of the estimated number of clusters, were overall higher for the \textit{walktrap} method (Figure \ref{fig:figure9}), while the mean F1 score favored both \textit{scran} methods (\textit{walktrap} and \textit{fast greedy}) over \textit{Seurat}. Finally, the ARI score at the true number of clusters, when available, showed similar performances. However, because \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified (Supplementary Figures 1 and 23), \textit{Seurat} could always be coerced into producing the right number of clusters. Instead, the number of clusters found by \textit{scran} was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Supplementary Figure 24), and as a result \textit{scran}-based clustering sometimes never produced a partitioning with the right number of clusters. This observation, along with \textit{scran}'s higher MI score, suggest that \textit{scran} sometimes simply divides a real subpopulation into two clusters (possibly tracking some unknown biological differences) rather than committing misclassification errors. Overall, the \textit{walktrap} method appeared superior to the \textit{fast greedy} algorithm and was generally less prone to misclassification than \textit{Seurat} clustering, although the latter offered more control over the resolution. A major difference between \textit{walktrap}-based clustering and \textit{Seurat} is the computing time (Figure \ref{fig:figure9}B). We found that using the Annoy approximation to the nearest neighbors somewhat reduced computing time at no apparent cost to accuracy. It nevertheless remained considerably slower than \textit{Seurat}. For all methods, some poorly distinguishable subpopulations from both the Zhengmix8eq and simMix1 datasets remained very inaccurately classified in regard to all metrics.

@@

To illustrate the applicability of \textit{pipeComp} to a completely different setting, we benchmarked the impact of surrogate variable analysis (SVA) -based methods on bulk RNAseq differential expression analysis (DEA). The rationale behind such approaches is that the identification of factors determining genome-wide variation (i.e. surrogate variables) and their inclusion in the generalized linear model for DEA might improve DEA, especially in the presence of sources of variation unrelated to the experimental variables. \textit{sva}\citep{leekCapturingHeterogeneityGene2007,leekSvaPackageRemoving2012} and \textit{RUVseq}\citep{rissoNormalizationRNAseq2014a} are two popular packages implementing methods to infer these variables. We therefore created a \texttt{PipelineDefinition} with three step: 1) feature filtering, 2) removal of unwanted variation (i.e. SVA), and 3) DEA using the most established methods (the package's `pipeComp\_dea' vignette details the creation process). We developed 3 benchmark datasets: 1) the \texttt{ipsc} data is a random selection of 10 vs 10 samples from \citet{carcamoOriveAnalysisTranscriptional2017}, which are very heterogeneous, and foldchanges were manually added to 300 genes; 2) the \texttt{seqc} dataset contains 5 vs 5 samples of two mixtures of the SEQC consortium with different ERCC spike-in mixes, and includes a partially correlated batch effect; 3) A **simulation** of 8 vs 8 samples with two technical vectors of variation (see methods for more details). As previously reported, we found that more stringent filtering improved DEA (Supplementary Figure 25A) and that the different DEA methods had similar performances (Figure \ref{fig:svadea}A and Supplementary Figure 25A). Including a SVA-based step increased sensitivity across all datasets, and particularly so in datasets with technical vectors of variation, and benefited all DEA methods in a similar fashion (Figure \ref{fig:svadea}A,C and Supplementary Figures 25-26). With the exception of the \texttt{RUVr} method, which led to a high False Discovery Rate (FDR), all other methods showed comparable gains in sensitivity while maintaining a decent error control (if a little above the nominal FDR), even when increasing the number of surrogate variables used (Figure \ref{fig:svadea}B-C). These results therefore suggest that these methods are robust and can be widely applied.

@@

@@

\begin{figure}
    \centering
    \ifshowfigs
    \includegraphics[width=\textwidth,keepaspectratio]{dataset_description}
    \fi
    \caption{\textbf{Overview of the benchmark datasets used.}}
    \label{fig:figure1}
\end{figure}

\begin{figure}
    \centering
    \ifshowfigs
    \includegraphics[width=\textwidth, keepaspectratio]{main_figures/pipeline_explanation.pdf}
    \fi
    \caption{\textbf{Overview of the \textit{pipeComp} framework and its application to a scRNAseq clustering pipeline. A:} The package is built around a \texttt{PipelineDefinition} class which defines a set of functions to be executed consecutively, as well as optional evaluation functions for each step. Each step can accept a number of parameters whose alternative values are provided as a list. All (or subsets of) combinations of parameters can then be simultaneously ran and evaluated using the \texttt{runPipeline} function (see Supplementary Methods for more details). \textbf{B:} Scheme representing the application of \textit{pipeComp} to evaluate a range of methods that are commonly used in scRNAseq studies and some of the metrics monitored at various steps.}
    \label{fig:figure2}
\end{figure}

\begin{figure}
    \centering
    \ifshowfigs
    \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/doublets.pdf}}
    \fi
    \caption{\textbf{Identification of doublet cells. A:} Receiver operating characteristic (ROC) curves of the tested doublet detection methods for three datasets with SNP-identified doublets. Dots indicate threshold determined by the true number of doublets. \textbf{B:} Running time of the different methods (\textit{DoubletFinder} failed on one of the datasets). \textbf{C:} Rate of misclassification of the cells identified by \textit{scDblFinder} as doublets (DBL) or singlets (SNG), across a large range of clustering analysis. Even in datasets which should not have neotypic doublets, the cells identified as such tend to be misclassified.    \label{fig:figure3}}
\end{figure}

>>>>

The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (Figure \ref{fig:explanation}A). In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics. Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments (or a desired subset), avoiding recomputing the same step twice without the need to save all alternative intermediates, and compiling evaluations (including run time) on the fly (see Supplementary methods for more detail). Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline. This is especially important because end-point metrics, such as the adjusted Rand index (ARI) \cite{HubertARI1985} for clustering, are not perfect. For example, although the meaning of an ARI score is independent of the number of true subpopulations \cite{steinleyProperties2004}, the number of clusters called is by far the most important determinant of the score: the farther it is from the actual number of subpopulations, the worse the ARI (Supplementary Figures 1-2). In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of clusters \cite{duoClustering2018}. While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown and tools that operate well in this optimal context might not necessarily be best overall. Clustering results are also very sensitive and might not always capture improvements in earlier steps of the pipeline. In addition, we wanted to assess whether the effect of a given parameter alteration is robust to changes in the rest of the pipeline. We therefore monitored several complementary metrics across multiple steps of the process. We proceeded in a step-wise fashion, first testing a large variety of parameters at the early steps of the pipeline along with only a set of mainstream options downstream, then selecting main alternatives and proceeding to a more detailed benchmark of the next step (Figure \ref{fig:explanation}B).

@@

While most methods accurately identified the doublets in the 3 cell lines dataset (mixology10x3cl), the other two datasets proved more difficult (Figure \ref{fig:doublets}A). \textit{scDblFinder} achieved comparable or better accuracy than top alternatives while being the fastest method (Figure \ref{fig:doublets}B). Across datasets, cells called as doublets tended to be classified in the wrong cluster more often than other cells (Figure \ref{fig:doublets}C). We also found that \textit{scDblFinder} improved the accuracy of the clustering across all benchmark datasets even when, by design, the data contained no heterotypic doublet (Figure \ref{fig:filtering}).

@@

We next investigated the impact of filtering according to six methods with varying stringency; three methods based on deviations to MADs with increasing levels of stringency (named \textit{lenient}, \textit{default}, \textit{stringent}, \textit{very stringent}) and two methods based on \textit{scater}'s \textit{runPCA} using all (\textit{pca.all}) or selected covariates (\textit{pca.sel}, see Methods for more details). An important risk of excluding cells on the basis of their distance from the whole distribution of cells on some properties (e.g., library size) is that these properties tend to have different distributions across subpopulations. As a result, thresholds in terms of number of MADs from the whole distribution can lead to strong biases against certain subpopulations (Figure \ref{fig:filtering}A). We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \ref{fig:filtering}B and Supplementary Figure 10). Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here. We therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm \cite{PapadimitriouHu1998} and monitored their mean F1 score. A first observation was that, although the stringent filtering tended to be associated with an increase in accuracy, it could also become deleterious. Most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias (Figure \ref{fig:filtering}B). Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) rather than on whole dataset resulted in nearly no cells being filtered out. This suggests that stringent filtering on the global population tends to discard cells of subpopulations with more extreme properties (e.g. high library size), rather than low-quality cells. On the opposite, too lenient filtering (or lack thereof) can lead to misclassified cells, as was for instance the case especially with the Zhengmix4 datsets (Figure \ref{fig:filtering}B). We found that a relatively mild filtering (`default' - see methods) often leads to high clustering performance (e.g. simMix and Zhengmix4 datasets) and always retain most of the cells in each subpopulation (e.g. Koh and Zhengmix8eq). This suggests that a mild ('default') filtering is a very good trade-off between clustering accuracy and subpopulations preservation.  Finally, the doublet removal step in conjunction with filtering  resulted in a net decrease in the proportion of excluded cells while retaining or improving accuracy not only in datasets that do include heterotypic doublets (mixology datasets), but also some others in which such doublets are unlikely (e.g. Zheng datasets). On the Smart-Seq and simulated datasets however, the removal of doublets had little to no impact on any of the two measures. Overall, we therefore recommend the use of doublet removal followed by our `default' filtering (see Methods) or a similar approach.

@@

An important motivation for the development of \textit{SCTransform} was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an inadequate normalization \cite{hafemeisterSCtransform2019}. However, as library size tends to vary across subpopulations, part of this effect can simply reflect biological differences. We therefore assessed to what extent the first principal component still retained a correlation with the library size and the number of detected features, removing the confounding covariation with the subpopulations (Figure \ref{fig:norm}A and Supplementary Figure 12). Scaling tended to remove much of the correlation with these features, and while most methods were able to remove most of the effect, the correlation was lowest with \textit{SCTransform}. Surprisingly, however, regressing out the covariates tended to increase the association with it.

We further evaluated normalization methods by investigating their impact on the separability of the subpopulations (Figure \ref{fig:norm}B-C). Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width \cite{RousseeuwSil1987} and mutual information (MI). We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations. In view of the subpopulation silhouette and ARI at true number of clusters, it seems that all methods perform worse on the three Zhengmix datasets, that are the datasets with the highest number of cells and the smaller feature detection rate. Scaling (e.g. `seurat (no scale)' vs `seurat') tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones and was generally, but not always, beneficial on the accuracy of the final clustering. Regressing out covariates systematically gave equal  or poorer performance on MI, ARI and ARI at true number of clusters, with a notable negative impact when \textit{Seurat}'s regressing was applied to the Koh and Kumar datasets. It is possible that the difference between \textit{Seurat} and \textit{SCTransform}'s regression originates from the use of Pearson's residuals performed during the second round regression of \textit{SCTransform}, which are then independent of sequencing depth. \textit{SCTransform} systematically outperformed other methods and, even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).\textit{scVI} (Supplementary Figure 13) performed poorly on most datasets in terms of both the subpopulation silhouette and clustering accuracy. For time and resource cost reasons, \textit{scVI} was run on a reduced set of pipelines and compared to \textit{SCTransform} (more details in the Methods section).

@@

Reasoning that a selection method should ideally select genes whose variability is higher between subpopulations than within, we first assessed to what extent each method selected genes with a high proportion of variance or deviance explained by the (real) subpopulations. As the proportion of variability in a gene attributable to subpopulations can be measured in various ways, we first compared three approaches: ANOVA on log-normalized count, ANOVA on \textit{SCTransform} normalization, and deviance explained. The ANOVAs performed on a standard \textit{Seurat} normalization and on \textit{SCTransform} data were highly correlated (Pearson correlation >0.9, Supplementary Figure 17A). These estimates were also in good agreement with the deviance explained, although lowly-expressed genes could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 17B-D). We therefore compared the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 18-19). We first focused on the first 1000 genes to highlight the differences between methods, although a higher number of selected genes decreased the differences between methods (Supplementary Figures 18-20). The standardized measures of variability were systematically worse than their non-standardized counterparts in selecting genes with a high proportion of variance explained by subpopulation. Regarding the percentage of deviance explained however, the standardized measures were often superior (Figure \ref{fig:selection}A and Supplementary Figures 18-19). \textit{Deviance} proved the method of choice to prioritize genes with a high variance explained by subpopulations (with mere expression level proving surprisingly performant) but did not perform so well to select genes with a high deviance explained.

We next evaluated how the use of different feature selection methods affected the clustering accuracy (Figure \ref{fig:selection}B). To validate the previous assay on the proportion of variance/deviance explained by real populations, we included genes that maximized these two latter measures. Interestingly, while these selections were on average the top-ranking methods, they were not systematically best for all datasets. The ARI at true number of clusters was near perfect (>0.94) among the smallest datasets but among bigger datasets (simMix and ZhengMix) the performance of the methods was lower and more variable (Figure \ref{fig:selection}B). Non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics. In general, we found deviance and unstandardized estimates of variance to provide the best results across datasets and normalization methods. Increasing the number of features selected also systematically led to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Supplementary Figure 20).

@@

We compared the various estimates of dimensionality in their ability to retrieve the intrinsic number of dimensions in a dataset, based on Seurat's weighted PCA space. As a first approximation of the true dimensionality, we computed the variance in each principal component that was explained by the subpopulations, which sharply decreased after the first few components in most datasets (Figure \ref{fig:dim}A). Figure \ref{fig:dim}B shows the difference between the dimension estimates of the above methods and that based on the subpopulations (i.e. from Figure \ref{fig:dim}A). Of note, the methods differ widely in view of their computing time (Figure \ref{fig:dim}B) and we saw no relationship between the accuracy of the estimates and the complexity of the method. Reasoning that over-estimating the number of dimensions is less problematic than under-estimating it, we kept the former methods for a full analysis of their impact on clustering (Figure \ref{fig:dim}C-D), when combined with \textit{SCTransform} or the standard \textit{Seurat} normalization. Although most methods performed well on the various clustering measures, the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}, using the 20 nearest neighbors, although estimates were relatively stable across decent values of \textit{k}; see Supplementary Figure 23) provided the dimensionality estimate that best separated the subpopulations (Figure \ref{fig:dim}C) and resulted in the best clustering accuracy (Figure \ref{fig:dim}D). This method systematically selected many more components than were associated with the subpopulations, suggesting that although these additional components appear individually uninformative, in combination they nevertheless contribute to classification. An implication is that the commonly-used elbow method most likely underestimates dimensionality.

@@

Since ARI is dominated by differences in the number of clusters (Supplementary Figures 1-2) and no single metric is perfect, we diversified them (Figure \ref{fig:clust}). MI has the virtue of not decreasing when a true subpopulation is split into two clusters, which is arguably less problematic (and might well reflect unknown biological subgroups), but as a consequence it can be biased towards methods producing higher resolution clustering (Supplementary Figure 2). Similarly, precision per true subpopulation is considerably more robust to differences in the number of clusters. We also tracked the mean F1 score and the ARI at the true number of clusters.

The MI score and minimum precision, which are largely independent of the estimated number of clusters, were overall higher for the \textit{walktrap} method (Figure \ref{fig:clust}), while the mean F1 score favored both \textit{scran} methods (\textit{walktrap} and \textit{fast greedy}) over \textit{Seurat}. Finally, the ARI score at the true number of clusters, when available, showed similar performances. However, because \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified (Supplementary Figures 1 and 23), \textit{Seurat} could always be coerced into producing the right number of clusters. Instead, the number of clusters found by \textit{scran} was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Supplementary Figure 24), and as a result \textit{scran}-based clustering sometimes never produced a partitioning with the right number of clusters. This observation, along with \textit{scran}'s higher MI score, suggest that \textit{scran} sometimes simply divides a real subpopulation into two clusters (possibly tracking some unknown biological differences) rather than committing misclassification errors. Overall, the \textit{walktrap} method appeared superior to the \textit{fast greedy} algorithm and was generally less prone to misclassification than \textit{Seurat} clustering, although the latter offered more control over the resolution. A major difference between \textit{walktrap}-based clustering and \textit{Seurat} is the computing time (Figure \ref{fig:clust}B). We found that using the Annoy approximation to the nearest neighbors somewhat reduced computing time at no apparent cost to accuracy. It nevertheless remained considerably slower than \textit{Seurat}. For all methods, some poorly distinguishable subpopulations from both the Zhengmix8eq and simMix1 datasets remained very inaccurately classified in regard to all metrics.

@@

To illustrate the applicability of \textit{pipeComp} to a completely different setting, we benchmarked the impact of surrogate variable analysis (SVA) -based methods on bulk RNAseq differential expression analysis (DEA). The rationale behind such approaches is that the identification of factors determining genome-wide variation (i.e. surrogate variables) and their inclusion in the generalized linear model for DEA might improve DEA, especially in the presence of sources of variation unrelated to the experimental variables. \textit{sva}\citep{leekCapturingHeterogeneityGene2007,leekSvaPackageRemoving2012} and \textit{RUVseq}\citep{rissoNormalizationRNAseq2014a} are two popular packages implementing methods to infer these variables. We therefore created a \texttt{PipelineDefinition} with three step: 1) feature filtering, 2) removal of unwanted variation (i.e. SVA), and 3) DEA using the most established methods (the package's `pipeComp\_dea' vignette details the creation process). We developed 3 benchmark datasets: 1) the \texttt{ipsc} data is a random selection of 10 vs 10 samples from \citet{carcamoOriveAnalysisTranscriptional2017}, which are very heterogeneous, and foldchanges were manually added to 300 genes; 2) the \texttt{seqc} dataset contains 5 vs 5 samples of two mixtures of the SEQC consortium with different ERCC spike-in mixes \citep{ComprehensiveAssessment2014}, and includes a partially correlated batch effect; 3) A **simulation** of 8 vs 8 samples with two technical vectors of variation (see methods for more details). As previously reported, we found that more stringent filtering improved DEA (Supplementary Figure 25A) and that the different DEA methods had similar performances (Figure \ref{fig:svadea}A and Supplementary Figure 25A). Including a SVA-based step increased sensitivity across all datasets, and particularly so in datasets with technical vectors of variation, and benefited all DEA methods in a similar fashion (Figure \ref{fig:svadea}A,C and Supplementary Figures 25-26). With the exception of the \texttt{RUVr} method, which led to a high False Discovery Rate (FDR), all other methods showed comparable gains in sensitivity while maintaining a decent error control (if a little above the nominal FDR), even when increasing the number of surrogate variables used (Figure \ref{fig:svadea}B-C). These results therefore suggest that these methods are robust and can be widely applied.

@@

The full process of creating the SVA-DEA \texttt{PipelineDefinition} is described step by step in the \href{https://github.com/plger/pipeComp/blob/master/vignettes/pipeComp_dea.Rmd}{pipeComp\_dea vignette}.

The three benchmark datasets are available in \url{https://github.com/markrobinsonuzh/scRNA\_pipelines\_paper/blob/master/svadea/datasets/}, along with all code used to generate and prepare them. Briefly:
\begin{itemize}
    \item For the \textit{ipsc} dataset, we took the quantification of the GSE79636 dataset (fairly heterogeneous) from the \textit{RNAontheBENCH} package \citep{germainRNAontheBENCHComputationalEmpirical2016}, and selected two random sets of 10 samples. No further variation was added, and log-foldchanges ranging from 0.25 to 1.25 were added to the counts of 300 random genes. All other genes were assumed to be negatives.
    \item For the \textit{seqc} dataset, we used data from the \textit{seqc} package \citep{ComprehensiveAssessment2014}. For both groups (mixtures C and D, which also contain two different spike-in mixes), we selected samples from the \texttt{ILM\_refseq\_gene\_AGR} and \texttt{ILM\_refseq\_gene\_CNL} batches, which have clear technical differences; we selected 5 samples per group with a weak correlation with the batch (3:2 vs 2:3). Since the true differences between mixtures are not entirely known, the analysis was performed on all genes but only the spike-ins were considered for benchmarking.
    \item For the \textit{simulation}, we simulated counts using the \textit{polyester} package \citep{frazeePolyesterSimulating2015} and the means/dispersions from the GSE79636 dataset (restricted to a single batch and biological group), with 500 DEGs (log-foldchanges ranging from 0.5 to 2) and two batch effects: 1) a technical batch partially correlated with the group (6:2) and affecting 1/3 of the genes, and 2) a linear vector of variation uncorrelated with the groups and affecting 1/3 of the genes.
\end{itemize}

@@

\begin{figure}
    \centering
    \ifshowfigs
    \includegraphics[width=\textwidth, keepaspectratio]{main_figures/pipeline_explanation.pdf}
    \fi
    \caption{\textbf{Overview of the \textit{pipeComp} framework and its application to a scRNAseq clustering pipeline. A:} The package is built around a \texttt{PipelineDefinition} class which defines a set of functions to be executed consecutively, as well as optional evaluation functions for each step. Each step can accept a number of parameters whose alternative values are provided as a list. All (or subsets of) combinations of parameters can then be simultaneously ran and evaluated using the \texttt{runPipeline} function (see Supplementary Methods for more details). \textbf{B:} Scheme representing the application of \textit{pipeComp} to evaluate a range of methods that are commonly used in scRNAseq studies and some of the metrics monitored at various steps.}
    \label{fig:explanation}
\end{figure}

\begin{figure}
    \centering
    \ifshowfigs
    \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/doublets.pdf}}
    \fi
    \caption{\textbf{Identification of doublet cells. A:} Receiver operating characteristic (ROC) curves of the tested doublet detection methods for three datasets with SNP-identified doublets. Dots indicate threshold determined by the true number of doublets. \textbf{B:} Running time of the different methods (\textit{DoubletFinder} failed on one of the datasets). \textbf{C:} Rate of misclassification of the cells identified by \textit{scDblFinder} as doublets (DBL) or singlets (SNG), across a large range of clustering analysis. Even in datasets which should not have neotypic doublets, the cells identified as such tend to be misclassified.    \label{fig:doublets}}
\end{figure}

\begin{figure}
    \centering
    \ifshowfigs
    \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/filtering.pdf}}
    \fi
    \caption{\textbf{Effect of filtering on cell subpopulation structure and clustering. A:} Filtering on the basis of distance to the whole distribution can lead to strong bias against certain subpopulations. The dashed line indicates a threshold of 2.5 median absolute deviations (MADs) from the median of the overall population. \textbf{B:} Relationship between the maximum subpopulation exclusion rate and the average clustering accuracy per subpopulation across various filtering strategies. Of note, doublet removal appears to be desirable even when, due to the design, there are no heterotypic doublets in the data. The PCA methods refer to multivariate outlier detected as implemented in \textit{scater} (see Methods for details).}
    \label{fig:filtering}
\end{figure}
